---
title: "Introduction to ppforest2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ppforest2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

ppforest2 builds oblique decision trees and random forests using projection pursuit. Instead of splitting on a single variable at each node, it finds a linear combination of variables that best separates the groups.

## Single tree

Train a projection-pursuit tree on the iris dataset:

```{r tree}
library(ppforest2)

tree <- pptr(Species ~ ., data = iris, seed = 0)
tree
```

The tree splits on linear projections of the features. Use `summary()` to see variable importance:

```{r tree-summary}
summary(tree)
```

Predict new observations:

```{r tree-predict}
preds <- predict(tree, iris)
table(predicted = preds, actual = iris$Species)
```

## Random forest

Train a forest of 100 trees, each considering 2 randomly chosen variables at each split:

```{r forest}
forest <- pprf(Species ~ ., data = iris, size = 100, n_vars = 2, seed = 0)
summary(forest)
```

The summary shows the OOB (out-of-bag) error estimate and three variable importance measures:

- **Projection**: average absolute projection coefficient across all splits
- **Weighted**: projection importance weighted by the number of observations routed through each node
- **Permuted**: decrease in OOB accuracy when each variable is permuted

Predict with class probabilities:

```{r forest-predict}
probs <- predict(forest, iris[1:5, ], type = "prob")
probs
```

## Visualization

ppforest2 provides several plot types via `plot()` (requires ggplot2):

```{r plot-structure, eval = requireNamespace("ggplot2", quietly = TRUE)}
# Tree structure with projected data histograms at each node
plot(tree, type = "structure")
```

```{r plot-importance, eval = requireNamespace("ggplot2", quietly = TRUE)}
# Variable importance bar chart
plot(tree, type = "importance")
```

```{r plot-projection, eval = requireNamespace("ggplot2", quietly = TRUE)}
# Data projected onto the first split's projection vector
plot(tree, type = "projection")
```

```{r plot-boundaries, eval = requireNamespace("ggplot2", quietly = TRUE)}
# Decision boundaries for two selected variables
plot(tree, type = "boundaries")
```

For forests, variable importance is the only global visualization available. Individual trees can be plotted by specifying a `tree_index`.

```{r plot-forest, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(forest)
```

```{r plot-forest-tree, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(forest, type = "structure", tree_index = 1)
```



## Regularization with PDA

Set `lambda > 0` to use Penalized Discriminant Analysis instead of LDA. This can help when features are highly correlated:

```{r pda}
tree_pda <- pptr(Species ~ ., data = iris, lambda = 0.5, seed = 0)
summary(tree_pda)
```

## Explicit strategy selection

For more control, you can pass strategy objects directly instead of using the shortcut parameters (`lambda`, `n_vars`, `p_vars`). This is equivalent but makes the strategy choice explicit:

```{r strategies}
# These two calls produce identical results:
forest_shortcut <- pprf(Species ~ ., data = iris, size = 10, lambda = 0.5, n_vars = 2, seed = 0)

forest_explicit <- pprf(Species ~ ., data = iris, size = 10, pp = pp_pda(0.5), vars = vars_uniform(n_vars = 2), seed = 0)

all.equal(predict(forest_shortcut, iris), predict(forest_explicit, iris))
```

Available strategy constructors:

- `pp_pda(lambda)` — PDA projection pursuit (`lambda = 0` for LDA)
- `vars_uniform(n_vars)` or `vars_uniform(p_vars)` — random variable selection
- `vars_all()` — use all variables (default for single trees)
- `cutpoint_mean_of_means()` — midpoint split rule (default)

Strategy objects can also be passed as engine arguments in parsnip:

```{r strategies-parsnip, eval = requireNamespace("parsnip", quietly = TRUE)}
library(parsnip)

spec <- pp_rand_forest(trees = 10) |>
  set_engine("ppforest2", pp = pp_pda(0.5), vars = vars_uniform(n_vars = 2)) |>
  set_mode("classification")

fit <- spec |> fit(Species ~ ., data = iris)
predict(fit, iris[1:5, ])
```

## Tidymodels integration

ppforest2 integrates with the tidymodels ecosystem via parsnip:

```{r parsnip, eval = requireNamespace("parsnip", quietly = TRUE)}
library(parsnip)

# Single tree
spec <- pp_tree(penalty = 0.5) |>
  set_engine("ppforest2") |>
  set_mode("classification")

fit <- spec |> fit(Species ~ ., data = iris)
predict(fit, iris[1:5, ])
```

```{r parsnip-forest, eval = requireNamespace("parsnip", quietly = TRUE)}
# Random forest
spec <- pp_rand_forest(trees = 50, mtry = 2, penalty = 0.5) |>
  set_engine("ppforest2") |>
  set_mode("classification")

fit <- spec |> fit(Species ~ ., data = iris)
predict(fit, iris[1:5, ], type = "prob")
```
