---
title: "Introduction to ppforest2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ppforest2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

ppforest2 builds oblique decision trees and random forests using projection pursuit. Instead of splitting on a single variable at each node, it finds a linear combination of variables that best separates the classes.

## Single tree

Train a projection-pursuit tree on the iris dataset:

```{r tree}
library(ppforest2)

tree <- pptr(Type ~ ., data = iris, seed = 42)
tree
```

The tree splits on linear projections of the features. Use `summary()` to see variable importance:

```{r tree-summary}
summary(tree)
```

Predict new observations:

```{r tree-predict}
preds <- predict(tree, iris)
table(predicted = preds, actual = iris$Type)
```

## Random forest

Train a forest of 100 trees, each considering 2 randomly chosen variables at each split:

```{r forest}
forest <- pprf(Type ~ ., data = iris, size = 100, n_vars = 2, seed = 42)
summary(forest)
```

The summary shows the OOB (out-of-bag) error estimate and three variable importance measures:

- **Projection**: average absolute projection coefficient across all splits
- **Weighted**: projection importance weighted by the number of observations routed through each node
- **Permuted**: decrease in OOB accuracy when each variable is permuted

Predict with class probabilities:

```{r forest-predict}
probs <- predict(forest, iris[1:5, ], type = "prob")
probs
```

## Visualization

ppforest2 provides several plot types via `plot()` (requires ggplot2):

```{r plot-structure, eval = requireNamespace("ggplot2", quietly = TRUE)}
# Tree structure with projected data histograms at each node
plot(tree, type = "structure")
```

```{r plot-importance, eval = requireNamespace("ggplot2", quietly = TRUE)}
# Variable importance bar chart
plot(tree, type = "importance")
```

```{r plot-projection, eval = requireNamespace("ggplot2", quietly = TRUE)}
# Data projected onto the first split's projection vector
plot(tree, type = "projection")
```

```{r plot-boundaries, eval = requireNamespace("ggplot2", quietly = TRUE)}
# Decision boundaries for two selected variables
plot(tree, type = "boundaries")
```

For forests, specify a `tree_index` to plot an individual tree:

```{r plot-forest-tree, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(forest, type = "structure", tree_index = 1)
```

## Regularization with PDA

Set `lambda > 0` to use Penalized Discriminant Analysis instead of LDA. This can help when features are highly correlated:

```{r pda}
tree_pda <- pptr(Type ~ ., data = iris, lambda = 0.5, seed = 42)
summary(tree_pda)
```

## Tidymodels integration

ppforest2 integrates with the tidymodels ecosystem via parsnip:

```{r parsnip, eval = requireNamespace("parsnip", quietly = TRUE)}
library(parsnip)

# Single tree
spec <- pp_tree(penalty = 0.5) |>
  set_engine("ppforest2") |>
  set_mode("classification")

fit <- spec |> fit(Type ~ ., data = iris)
predict(fit, iris[1:5, ])
```

```{r parsnip-forest, eval = requireNamespace("parsnip", quietly = TRUE)}
# Random forest
spec <- pp_rand_forest(trees = 50, mtry = 2, penalty = 0.5) |>
  set_engine("ppforest2") |>
  set_mode("classification")

fit <- spec |> fit(Type ~ ., data = iris)
predict(fit, iris[1:5, ], type = "prob")
```
